---
title: "Wright Maps"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Wright Maps}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(eatPrep)
```

`plotWrightMap()` places a person distribution and item names on a shared
vertical metric. Higher scores and harder items appear towards the top. A
continuous line separates the two sides; item names on the same display stage
are separated by ` | `. The default uses a soft teal distribution on a white
background with subtle guides. Both portrait and landscape output are supported.

Supply item difficulties and plausible values (PVs) that already refer to the
same dimension and metric. If identifiers repeat, the function checks the
`category` column (or the column specified by `category_col`). Identifier and
category together must be unique. Only repeated identifiers get a suffix, such
as `Item_03_cat1`; unique identifiers keep their names, and their categories may
be missing. Already unique names are also accepted. No threshold calculation is
performed.

```{r data}
set.seed(42)
persons <- data.frame(
  id = 1:500,
  PV1 = rnorm(500), PV2 = rnorm(500), PV3 = rnorm(500),
  weight = runif(500, 0.5, 2)
)
items <- data.frame(
  item = c("Item_01", "Item_02", "Item_03", "Item_03",
           "Item_04", "Item_05", "Item_06", "Item_07"),
  category = c(NA, NA, 1, 2, NA, NA, NA, NA),
  difficulty = c(-1.7, -1.1, -0.4, 0.35, 0.45, 0.8, 1.2, 1.75)
)
```

The default is a density curve. Each PV contributes a separate weighted density,
evaluated on a common grid with a common bandwidth; these curves are averaged.
The function never substitutes respondent-level PV means for the PV distribution.

```{r density, fig.width=7, fig.height=6}
p <- plotWrightMap(
  items, persons, pv_cols = paste0("PV", 1:3), weight_col = "weight",
  item_step = 0.25, score_label = "Ability / difficulty"
)
p
```

`item_step` controls display precision. Here each item is assigned to the nearest
multiple of 0.25, so its displayed position differs by at most 0.125 from the
supplied parameter. Exact halfway values are assigned upwards. The default
chooses a pretty step targeting about 30 intervals across the combined observed
PV and item range. `item_step = 0` retains exact positions. Item names retain
input order within each stage.

The original parameters and computed distribution remain available:

```{r data-access}
attr(p, "wright_data")$items
head(attr(p, "wright_data")$population)
```

An optional histogram uses common left-closed, right-open bins for all PVs and
averages weighted relative frequencies across PVs. Both histogram and density
widths are scaled to show the distribution's shape, rather than person counts.

```{r histogram, fig.width=8, fig.height=5}
plotWrightMap(
  items, persons, pv_cols = paste0("PV", 1:3), weight_col = "weight",
  person_geom = "histogram", binwidth = 0.25, item_step = 0.25
)
```

Wide and long PV tables use the same arguments as `plotPopulationCuts()`:

```{r long, eval=FALSE}
long <- tidyr::pivot_longer(
  persons, c("PV1", "PV2", "PV3"), names_to = "pv", values_to = "score"
)
plotWrightMap(
  items, long, respondent_id_col = "id", pv_id_col = "pv",
  pv_value_col = "score", weight_col = "weight"
)
```

Missing PVs are rejected by default. `pv_missing = "drop"` omits them with a
warning and renormalizes weights within each PV. At least two observations with
positive weights must remain in every PV.

Use `person_prop` to set the approximate share of width for the person side,
`item_size` for the maximum item font size, and `density_bw` or `density_adjust`
for smoothing. Long rows wrap between item names, with a hanging indent for
continuation lines and a bracket grouping the block. Blocks can move slightly
to avoid overlap; leaders connect them to exact stage markers on the divider.
Text shrinks only when all blocks cannot fit, or a single name is too wide.
With many items, increase the output width or height, decrease `person_prop`, or
adjust `item_step` to keep text readable. `score_limits` zooms the display
without changing the estimates or stages.

```{r crowded, fig.width=9, fig.height=7}
crowded_items <- setNames(seq(-2, 2, length.out = 80),
                          paste0("Item_", sprintf("%02d", 1:80)))
plotWrightMap(crowded_items, persons, pv_cols = paste0("PV", 1:3),
              item_step = 0.5)
```

Colours are configurable through `person_fill`, `person_colour`, `item_colour`,
and `line_colour`. For a neutral appearance, for example:

```{r neutral, eval=FALSE}
plotWrightMap(items, persons, pv_cols = paste0("PV", 1:3),
              person_fill = "grey92", person_colour = "grey30",
              item_colour = "grey20", line_colour = "grey65")
```

Multiple populations can be overlaid on the person side using `population_col`,
with the same grouping and colour conventions as `plotPopulationCuts()`.
Distributions are computed separately within populations and PVs, with a common
bandwidth and score grid (or common histogram bins). A shared horizontal scaling
factor makes shapes comparable; widths do not represent population sample sizes.
Respondent IDs may recur across populations. Items and mean cuts remain common.

```{r populations, fig.width=9, fig.height=6}
group_a <- transform(persons, population = "Group A")
group_b <- transform(persons, population = "Group B", PV1 = PV1 + 0.8,
                     PV2 = PV2 + 0.8, PV3 = PV3 + 0.8)
both <- rbind(group_a, group_b)
plotWrightMap(items, both, pv_cols = paste0("PV", 1:3), weight_col = "weight",
              respondent_id_col = "id", population_col = "population",
              population_colors = c("Group A" = "#327D83", "Group B" = "#8C6BB1"),
              cuts = c(cut12 = -0.8, cut23 = 0.3, cut34 = 1.2),
              cut_value_digits = 1)
```

Grouped fills are transparent by default, with opaque outlines and a legend.
Use `population_alpha` to change fill opacity. `population_colors` sets both
fill and outline colours. The returned distribution table includes a
`population` column. The same grouping also works with `person_geom = "histogram"`
and with long-format PV input.

Mean cuts appear as horizontal lines through the person densities (or
histograms), ending at the divider before the item names. All populations share
the same lines at the exact cut scores. The right side remains fully available
for item names. Pass the same
numeric cut vector used by `plotPopulationCuts()`, a one-row `cuts_summary`
table, or a complete `computeCutsIDM()` result. For an IDM result, only
`cuts_summary` is used, corresponding to `cut_selection = "mean"` in
`plotPopulationCutsIDM()`.

```{r mean-cuts, fig.width=9, fig.height=6}
plotWrightMap(items, persons, pv_cols = paste0("PV", 1:3),
              weight_col = "weight", item_step = 0.25,
              cuts = c(cut12 = -0.8, cut23 = 0.3, cut34 = 1.2),
              cut_value_digits = 1)
```

```{r idm-cuts, eval=FALSE}
plotWrightMap(items, persons, pv_cols = paste0("PV", 1:3), cuts = idm_result)
# Equivalently: cuts = idm_result$cuts_summary
```

Use `cut_labels` to override names, `show_cut_values = FALSE` to show names
without numbers, and `cut_colour` / `cut_value_size` for styling. Equal cuts
share a line; nearby labels avoid overlap and remain connected to their lines.
Labels sit on the person side with white backgrounds to remain readable over
the distributions.
Display rounding does not affect the exact values, which remain accessible as
`attr(plot, "wright_data")$cuts`. Adding cuts does not change the estimated
population distribution or item stages. Cuts must already share the score
metric; missing mean cuts must be resolved before plotting.

The return value is a `ggplot` object. For example:

```{r save, eval=FALSE}
p + ggplot2::labs(title = "Wright map")
ggplot2::ggsave("wright-map.pdf", p, width = 8, height = 6)
```
